A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2022-10-31 Papers/Paper/manuscript.Rmd
Roger's edits

For instance, competing theories about why certain types of object-extracted relative clauses, like *the lawyer that the banker irritated*, are harder to understand than the corresponding subject-extracted relative clauses, like *the lawyer that irritated the banker*, make different predictions about which words are the loci of the overall difficulty and slower RTs associated with object relatives  [@traxlerProcessingSubjectObject2002]. Dependency-locality theory [@grodnerConsequencesSerialNature2005] predicts that the locus of difficulty on object relatives is on the verb of the embedded clause (*irritated* in *the lawyer that the banker irritated*). In contrast, surprisal theory predicts the locus of difficulty will be on the article *the* at the start of the embedded clause [@staubEyeMovementsProcessing2010]. RT measures can potentially also inform theories about the time course of processing (i.e. which steps are parallel versus serial, @bartekSearchOnlineLocality2011) or the functional form of relationships between word characteristics and processing time [@smithEffectWordPredictability2013].

@@

The role of word frequency is worth special note. Although the facilitative effect of word frequency on reading measures has been known for decades, it is theoretically unclear why this effect should exist in any model that is sensitive to conditional word probability:

@smith-levy:2008 noted this puzzle and posited that the role of frequency could diminish or disappear as measurements of surprisal improve (as they have over the past decade), or alternatively that the effects of frequency (in addition to those of surprisal) may be the result of quick, heuristic responses to the words based on the more easily available unigram frequency before more fine-grained context-sensitive surprisal is available. An alternative is that frequency effects don't actually exist, but are an artifact of how RT data is analysed [@shainLargescaleStudyEffects2019].

@@

  1. Do participants engage with these longer passages successfully with the A-maze task?
  2. Is A-maze as powerful and reliable a method as SPR for these longer passages?
  3. What is the functional form between word surprisal and RT for the A-maze task?
  4. Does A-maze have less spillover than SPR?
  5. What types of context-driven expectations, as operationalized in competing computational language models, are deployed to determine RTs in A-maze and SPR?

@@

## How do A-maze and SPR compare in reliability?

@@

Our second question was whether A-maze is reliable. To assess reliability, we conducted split-half comparisons looking at the correlations between and within SPR and A-maze. If the methods picked up on the same effects, we would expect them to be correlated, with sentences that took longer to read in one method also taking longer in the other. We calculated the average RT at the sentence level to reduce variability from spillover patterns. The correlation between Maze and SPR was `r round(spr_maze,2)`, compared to `r round(spr_spr,2)` within SPR and `r round(maze_maze,2)` within Maze. See Figure \@ref(fig:scatter) for a visual comparison of overall Maze versus SPR RTs.  SPR data is about as correlated with Maze as with another sample of SPR data which provides some evidence that Maze and SPR are measuring the same effects. The superior within-method split-half correlation we see for Maze relative to SPR, despite the smaller number of participants, suggests that it is the more powerful of the two methods (higher signal-to-noise ratio), consistent with the findings of @boyceMazeMadeEasy2020 for factorial experimental designs with isolated-sentence presentation.

@@

The Maze linear model effects are shown in Figure \@ref(fig:coeffs-maze) (See also Appendix B for a table of effects). Across all models, there were consistent large effects of length and surprisal at the current word, but minimal effects of frequency. This lack of frequency effects is inconsistent with the results usually reported for SPR and eye-tracking. There was a small interaction between surprisal and length at the current word.

@@

We followed the same process for the SPR data with results shown in Table \@ref(tab:spr-compare).  For SPR, GPT-2 and 5-gram models contain some value over each other model, which is less clear for Transformer-XL and GRNN. In terms of log likelihoods, we find that GPT-2 is better than 5-gram is better than GRNN is better than Transformer-XL, although differences are small. The relatively good fit of 5-gram models to SPR data compared with neural models matches results from @huSystematicAssessmentSyntactic2020 and @wilcoxPredictivePowerNeural2020, and contrasts with the Maze results, where the 5-gram model had the worst fit and did not provide additional predictive value over the other models. While the nature of the generalizations made by these neural network-based models are not fully understood, controlled tests have suggested that their next-word predictions often reflect deeper features of linguistic structure [@warstadt-etal:2020-BLiMP; @huSystematicAssessmentSyntactic2020], such as subject--verb agreement [@marvin-linzen:2018-targeted] and wh-dependencies [@wilcox-etal:2022-using-computational-models], and are sensitive over longer context windows, than n-gram models. The fact that the neural language models, dominate the 5-gram models for Maze but not SPR thus suggests that Maze RTs may be more sensitive than SPR RTs to richer language structure-related processes during real-time comprehension.

@@

Second, we took advantage of the pre-existing SPR corpus on Natural Stories to compare the RT profiles between Maze and SPR.  Maze and SPR pick up on similar features in words, as shown by the high correlations between Maze and SPR RTs on the sentence level. The correlation within Maze is higher than Maze to SPR correlation or SPR-SPR correlations which is evidence that Maze is less noisy that SPR and may pick up on deeper linguistic features than SPR.

@@

Fourth, we compared the spillover profiles between Maze and SPR. For Maze, we found large effects of the current word's surprisal and length, which dwarfed any spillover effects from previous word predictors. In contrast, for SPR, we found effects of roughly equal sizes from the current and previous words. ^[Furthermore, the typical spillover profile for SPR data may be worse than suggested by the Natural Stories corpus SPR data: for example, @smithEffectWordPredictability2013 found that most of a word's surprisal effect showed up only one to two words downstream.] Overall, Maze is a slower task than SPR, but it also has much larger effects of length and surprisal, perhaps due to requiring more focus, and thus generating less noisy data. 
We do not find frequency effects on the Maze data, but we do on the SPR data. This could be explained if frequency effects are a first rough approximation of in-context predictability, before the fuller context-sensitive surprisal information is available. In this case, faster methods like eye-tracking and SPR would show frequency effects (in addition to surprisal), but slower methods like Maze would not as the additional demands slow down the response, allowing more contextual information to be used. While this is a difference between Maze and other incremental processing methods, we do not consider it a flaw, since there is not strong theoretical justification for the presence of frequency effects (in addition to surprisal effects) in language processing. These differences do point to the importance of understanding task demands of different incremental processing methods.

Lastly, we examined how different language models fare at predicting human RT data. We found that overall, the models were more predictive of the A-maze data than SPR data; however, the hierarchy of the model's predictive performance also differed between the A-maze and SPR datasets. This difference suggests that how well a language model predicts human RTs may depend on task. Maze RTs were by far best predicted by neural network language models, whereas SPR RTs were predicted nearly as well by 5-gram models. Our understanding of the linguistic generalization capabilities and performance of these neural network models is still limited, but controlled tests in the NLP literature that analyze their behavior on classic psycholinguistics paradigms [@linzen-etal:2016tacl; @futrell-etal:2019-neural-language-models] suggest more human-like performance than n-gram models are capable of. These findings further add to the evidence that the Maze task is favorable for RT-based investigations of underlying linguistic processing in the human mind. More broadly, further comparisons between different processing methods on the same materials could be useful for a deeper understanding of how task demands influence language processing [ex. @bartekSearchOnlineLocality2011].

Overall, A-maze has excellent localization, although some models showed small but statistically reliable effects of the past word. On the whole, our results support the idea that Maze forces language processing to be close to word-by-word, and thus the Maze task can be used under the assumption that the RT of a word primarily reflects its own properties and not those of earlier words. Correlations between Maze and SPR indicate that Maze is picking up on similar patterns, but with less noise, than SPR.

>>>>

For instance, competing theories about why certain types of object-extracted relative clauses, like *the lawyer that the banker irritated*, are harder to understand than the corresponding subject-extracted relative clauses, like *the lawyer that irritated the banker*, make different predictions about which words are the loci of the overall difficulty and slower RTs associated with object relatives  [@traxlerProcessingSubjectObject2002]. Dependency-locality theory [@grodnerConsequencesSerialNature2005] predicts that the locus of difficulty on object relatives is on the verb of the embedded clause (*irritated* in *the lawyer that the banker irritated*). In contrast, surprisal theory predicts the locus of difficulty will be on the article *the* at the start of the embedded clause [@staubEyeMovementsProcessing2010; @vaniUsingInterpolatedMaze2021]. RT measures can potentially also inform theories about the time course of processing (i.e. which steps are parallel versus serial, @bartekSearchOnlineLocality2011) or the functional form of relationships between word characteristics and processing time [@smithEffectWordPredictability2013].

@@

The role of word frequency is worth special note. Although the facilitative effect of word frequency on reading measures has been known for decades, this effect remains a theoretical puzzle. Sensitivity to a word's probability in context can be derived from any of a number of optimization principles [@smithEffectWordPredictability2013; @levy:2013sentenceProcessing], and word frequency is a context-insensitive estimate of word probability. However, it is not straightforward why this effect should exist in models in which a role is played by *contextually-conditioned* word probability, which reading and other language processing measures are known to be sensitive to: conditioning on context should render the context-insensitive probability estimate irrelevant for optimal processing. @smith-levy:2008 noted this puzzle and posited two alternative explanations. First is estimation error of conditional word probability in computationally implemented language models used at the time, in which case the role of frequency could diminish or disappear as measurements of surprisal improve (as they have over the past decade); consistent with this hypothesis is the recent report of @shainLargescaleStudyEffects2019 finding no word frequency effects in reading using more recent language models. Second, the effects of frequency (in addition to those of surprisal) could be the result of quick, heuristic responses to the words based on the more easily available unigram frequency before more fine-grained context-sensitive surprisal becomes available within real-time language processing mechanisms.

@@

  1. Do participants engage with these longer passages successfully with the A-maze task?
  2. Is A-maze as sensitive a method as SPR for these longer passages?
  3. What is the functional form between word surprisal and RT for the A-maze task?
  4. Does A-maze have less spillover than SPR?
  5. What types of context-driven expectations, as operationalized in competing computational language models, are deployed to determine RTs in A-maze and SPR?

@@

## How do A-maze and SPR compare in sensitivity?

@@

Our second question was a comparison of the sensitivity (the signal-to-noise ratio) of A-maze and SPR. To assess sensitivity, we conducted split-half comparisons looking at the correlations between and within SPR and A-maze. If the methods picked up on the same effects, we would expect them to be correlated, with sentences that took longer to read in one method also taking longer in the other. We calculated the average RT at the sentence level to reduce variability from spillover patterns. The correlation between Maze and SPR was `r round(spr_maze,2)`, compared to `r round(spr_spr,2)` within SPR and `r round(maze_maze,2)` within Maze. See Figure \@ref(fig:scatter) for a visual comparison of overall Maze versus SPR RTs.  SPR data is about as correlated with Maze as with another sample of SPR data which provides some evidence that Maze and SPR are measuring the same effects. The superior within-method split-half correlation we see for Maze relative to SPR, despite the smaller number of participants, suggests that it is the more sensitive of the two methods (higher signal-to-noise ratio), consistent with the findings of @boyceMazeMadeEasy2020 for factorial experimental designs with isolated-sentence presentation.

@@

The Maze linear model effects are shown in Figure \@ref(fig:coeffs-maze) (See also Appendix B for a table of effects). Across all models, there were consistent large effects of length and surprisal at the current word, but minimal effects of frequency. This lack of frequency effects differs from the results usually reported for SPR and eye-tracking (though see @shainLargescaleStudyEffects2019). There was a small interaction between surprisal and length at the current word.

@@

We followed the same process for the SPR data with results shown in Table \@ref(tab:spr-compare).  For SPR, GPT-2 and 5-gram models contain some value over each other model, which is less clear for Transformer-XL and GRNN. In terms of log likelihoods, we find that GPT-2 is better than 5-gram is better than GRNN is better than Transformer-XL, although differences are small. The relatively good fit of 5-gram models to SPR data compared with neural models matches results from @huSystematicAssessmentSyntactic2020 and @wilcoxPredictivePowerNeural2020, and contrasts with the Maze results, where the 5-gram model had the worst fit and did not provide additional predictive value over the other models. While the nature of the generalizations made by these neural network-based models are not fully understood, controlled tests have suggested that their next-word predictions often reflect deeper features of linguistic structure [@warstadt-etal:2020-BLiMP; @huSystematicAssessmentSyntactic2020], such as subject--verb agreement [@marvin-linzen:2018-targeted] and wh-dependencies [@wilcox-etal:2022-using-computational-models], and are sensitive over longer context windows, than n-gram models. The fact that the neural language models dominate the 5-gram models for Maze but not SPR thus suggests that Maze RTs may be more sensitive than SPR RTs to richer language structure-related processes during real-time comprehension.

@@

Second, we took advantage of the pre-existing SPR corpus on Natural Stories to compare the RT profiles between Maze and SPR.  Maze and SPR pick up on similar features in words, as shown by the high correlations between Maze and SPR RTs on the sentence level. The correlation within Maze is higher than Maze to SPR correlation or SPR-SPR correlations, which is evidence that Maze is less noisy than SPR.

@@

Fourth, we compared the spillover profiles between Maze and SPR. For Maze, we found large effects of the current word's surprisal and length, which dwarfed any spillover effects from previous word predictors. In contrast, for SPR, we found effects of roughly equal sizes from the current and previous words. ^[Furthermore, the typical spillover profile for SPR data may be worse than suggested by the Natural Stories corpus SPR data: for example, @smithEffectWordPredictability2013 found that most of a word's surprisal effect showed up only one to two words downstream.] Overall, Maze is a slower task than SPR, but it also has much larger effects of length and surprisal, perhaps due to requiring more focus, and thus generating less noisy data. 
We do not find frequency effects on the Maze data, but we do on the SPR data. This could be explained if frequency effects are a first rough approximation of in-context predictability, before the fuller context-sensitive surprisal information is available. In this case, faster methods like eye-tracking and SPR would show frequency effects (in addition to surprisal), but slower methods like Maze would not as the additional demands slow down the response, allowing more contextual information to be used. While this is a difference between Maze and other incremental processing methods, we do not consider it a flaw for Maze---indeed, for researchers interested in focusing on context-contingent language processing, it may suggest an advantage for the Maze task. Regardless, these differences highlight the importance of understanding task demands of different incremental processing methods.

Lastly, we examined how different language models fare at predicting human RT data. We found that overall, the models were more predictive of the A-maze data than SPR data; however, the hierarchy of the model's predictive performance also differed between the A-maze and SPR datasets. This difference suggests that how well a language model predicts human RTs may depend on task. Maze RTs were by far best predicted by neural network language models, whereas SPR RTs were predicted nearly as well by 5-gram models. Our understanding of the linguistic generalization capabilities and performance of these neural network models is still limited, and there are cases where they are known to make more superficial, non-human-like generalizations [@mccoy-etal-2019-right; @chaves-2020-dont], but controlled tests in the NLP literature that analyze their behavior on classic psycholinguistics paradigms [@linzen-etal:2016tacl; @futrell-etal:2019-neural-language-models; @warstadt-etal:2020-BLiMP; @wilcox-etal:2022-using-computational-models] suggest more human-like performance than n-gram models are capable of. These findings further add to the evidence that the Maze task is favorable for RT-based investigations of underlying linguistic processing in the human mind. More broadly, further comparisons between different processing methods on the same materials could be useful for a deeper understanding of how task demands influence language processing [ex. @bartekSearchOnlineLocality2011].

Overall, A-maze has excellent localization, although some models showed small but statistically significant effects of the past word. On the whole, our results support the idea that Maze forces language processing to be close to word-by-word, and thus the Maze task can be used under the assumption that the RT of a word primarily reflects its own properties and not those of earlier words. Correlation analysis between Maze and SPR suggest that Maze is picking up on many of the same patterns as does SPR, but with less noise.
